# Reproducible Research: Peer Assessment 1

```{r echo=FALSE}
setwd('/Users/adam/Documents/code/datasciencecoursera/reproducible_research/RepData_PeerAssessment1')
```

## Loading and preprocessing the data
```{r}
activity <- read.csv('activity.csv', header=TRUE)
no_missing <- activity[!is.na(activity$steps),]
```


## What is mean total number of steps taken per day?
```{r}
total_by_day <- by(no_missing$steps, no_missing$date, FUN=sum)
hist(total_by_day, main='Total number of steps taken each day', xlab='Total steps per day', col='lightblue')
```

Mean and median by day:
```{r results='asis'}
library(xtable)
summaries <- merge(as.data.frame(as.matrix(by(no_missing$steps, no_missing$date, FUN=mean))), 
                   as.data.frame(as.matrix(by(no_missing$steps, no_missing$date, FUN=median))), 
                   by="row.names", all=TRUE)
colnames(summaries) <- c("Date", "Mean", "Median")
xt <- xtable(summaries)
print(xt, type="html")
```


## What is the average daily activity pattern?
```{r}
intervals <- as.data.frame(as.matrix(by(no_missing$steps, no_missing$interval, FUN=mean)))
intervals$interval <- rownames(intervals)
plot(rownames(intervals), intervals$V1, main='Average number of steps taken by interval', xlab='5-Minute Interval', ylab='Average Steps', type='l', col='darkblue')
```

Maximum average number of steps:
```{r}
intervals[intervals$V1 == max(intervals$V1), 2]
```

## Imputing missing values
1. The total number of missing values in the dataset
```{r}
length(activity[is.na(activity$steps), 1])
```

2. Devise a strategy for filling in all of the missing values in the dataset

The median value for the interval with the missing value will be used.  
Using the mean of the interval has no effect on the data.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
intervals_median <- as.data.frame(as.matrix(by(no_missing$steps, no_missing$interval, FUN=median)))
intervals_median$interval <- rownames(intervals_median)

fillMissing <- function(row) {
    if(is.na(row$steps)) {
        intervals_median[intervals_median$interval == row$interval, "V1"]   
    } else {
        row$steps
    }
}

imputed <- cbind(activity)
imputed$steps <- as.numeric(by(imputed, 1:nrow(activity), FUN=fillMissing))
```

4. The mean and median total number of steps taken per day
```{r}
total_by_day_filled <- by(imputed$steps, imputed$date, FUN=sum)
hist(total_by_day_filled, main='Total number of steps taken each day', xlab='Total steps per day', col='lightblue')
```

   Mean and median by day:
```{r results='asis'}
summaries_filled <- merge(as.data.frame(as.matrix(by(imputed$steps, imputed$date, FUN=mean))), 
                          as.data.frame(as.matrix(by(imputed$steps, imputed$date, FUN=median))), 
                          by="row.names", all=TRUE)
colnames(summaries_filled) <- c("Date", "Mean", "Median")
xt <- xtable(summaries_filled)
print(xt, type="html")
```

   Do these values differ from the estimates from the first part of the assignment? 
```{r}
print(summaries_filled$Mean - summaries$Mean)
print(summaries_filled$Median - summaries$Median)
```
 
   What is the impact of imputing missing data on the estimates of the total daily number of steps?

There is no impact on the data at a day-by-day level.  
As shown in the histogram above, there is an impact on the first bin, which includes 0 steps per day.
Since the median value was used to impute the missing values for each interval, in all cases it appears the imputed value was zero.
This changes the histogram but has no net effect on the data at a day-by-day level.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
imputed$dow <- weekdays(as.Date(imputed$date, '%Y-%m-%d'), FALSE)
setDow <- function(row) {
    if(row$dow == 'Saturday') {
        'weekend'   
    } else if (row$dow == 'Sunday') {
        'weekend'
    } else {
        'weekday'
    }
}
imputed$daytype <- as.factor(by(imputed, 1:nrow(imputed), FUN=setDow))
```

```{r}
weekday_avg <- as.data.frame(as.matrix(by(imputed[imputed$daytype=='weekday', "steps"], imputed[imputed$daytype=='weekday', "interval"], FUN=mean)))

weekend_avg <- as.data.frame(as.matrix(by(imputed[imputed$daytype=='weekend', "steps"], imputed[imputed$daytype=='weekend', "interval"], FUN=mean)))

par(mfrow=c(2,1), mar=c(4,3,3,2))
plot(rownames(weekend_avg), weekend_avg$V1, type='l', main='weekend', xlab='Interval', ylab='Average number of steps')
plot(rownames(weekday_avg), weekday_avg$V1, type='l', main='weekday', xlab='Interval', ylab='Average number of steps')
```
